Stability and Performance of Coalitions of Prosumers Through Diversification in the Smart Grid

### 2015-11-02 Article.tex
figures sizes
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section I: Introduction
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

Designing stable power systems is a classical engineering challenge since blackouts can have catastrophic consequences. One obvious condition for stability revolves around sustaining at any time a power generation that meets the demand. If one is larger than the other, the system deviates from its synchronous state. If nothing is done to return to the synchronized equilibrium, this can lead to catastrophic cascades of failures \cite{Brummitt2012} \cite{Wang2009}. With traditional power plants based on fossil energies, the production can be scheduled in order to sustain the predicted consumption. Deviations from this schedule can then be supported almost on real time by using fast response power plants, interconnections with border countries, as well as regulator entities. Even if most individual consumers may not realize it, the electricity prices are not constant rates and evolve with the production/consumption conditions on the grid. It is often assumed that the granularity of these prices is meant to increase in the future as a way to pass on the production conditions on the end users \cite{Jiang2014}. The pricing of electricity and the necessity for the grid operator to have different emergency reserves lead to an economy setting for electricity, where operations and various kind of contracts are decided on a market \cite{Europe}. Such a market environment clearly necessitates communication in order to monitor and obtain information from the grid as well as exchange between participants. These kind of settings are already being used at the transport level of current power grid with large renewable plants in the production portfolio. Using electricity markets appears thus as a way to manage the reliability of the whole system, especially when the use of renewables is important.

The smart grid vision however goes further and revolutionizes this top-down centralized architecture by assuming that bidirectional power flows are allowed. This would change completely the nature of the traditional grid since production could be located down to the very end of the distribution networks. Nodes that were simply pure loads yesterday could behave tomorrow as generators or loads \cite{Ramchurn}. On the other hand, the user of renewables in the production is continuously growing, and is expected to become majority in a near future. These plants completely rely on the presence and the intensity of their respective resources (wind, sun, tiles...). Balancing production and demand in such a scenario appears much more challenging and it seems inevitable that the grid should modernize its infrastructures to sustain this transition \cite{Europe}. The key to drive safely such a system is assumed to be information, and more precisely, the capacity to measure, communicate, and analyze data on real time.

In this paper, we focus on these "\textit{nodes}" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "\textit{prosumer}" \cite{Rathnayaka2012} (and will be called accordingly in this paper). A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation happen.

@@

@@

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar radiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix \ref{appendixB} and in \cite{Lydia2014}.

@@

\begin{figure}
\includegraphics[scale=.48]{gaussian_P_star}
\caption{{\footnotesize $ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for Gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horizontal dashed line).} }
\label{fig:Gaussian}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{mean_field}
\caption{{\footnotesize Gaussian mean approximation. Subplot \textbf{a} shows how the parameter $ \alpha $ of the utility function should be chosen in function of the mean size of the coalitions (see equation \ref{eq:alpha_star}). Subplot \textbf{b} displays the corresponding utility functions for different values of $ \alpha $. Blue curve with diamonds favors small coalitions of 2 agents while the green one with triangles favors 5 agents coalitions. Finally, the red curve with squares has an optimal size of 15 agents.} }
\label{fig:mean_approx}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{cliques_histo}
\caption{{\footnotesize Histograms of utility values for coalitions of size 3. Red bars stand for cliques in the decorrelation graph, and blue bars for all the other possible triplets. Cliques tend to exhibit higher utilities than random coalitions.}}
\label{fig:histo_cliques}
\end{figure}

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, we do not have any technical constraints on coalition sizes even if we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents.
\begin{figure}
\includegraphics[scale=.48]{real_utility_2}
\caption{{\footnotesize Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to eq. \ref{eq:alpha_star} in order to favor 40 agents coalitions.}}
\label{fig:real_utility2}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figures/coalitions/coalitions6}
\caption{{\footnotesize Coalitions formed in the (contract value, volatility) space. The color map indicates the underlying quality. The closer to red the better (high contract values with small volatility). On the opposite, blue areas show poor quality (small contract values with high volatility). Blue dots stand for the decorrelated coalitions that we formed while green squares show correlated coalitions. The smaller yellow markers stand for the gravity centers of the coalition structures. The black dotted line shows how contract values and volatility evolve  when the size of the coalitions increases (a subset of the points are labeled by the size of the coalition they represent). Each point is the average over $ 10^{5} $ unconstrained draws of a random coalition. We also show the distribution of the gravity centers of random coalition structures with red ellipses (center is the mean, and each ellipse corresponds to one  standard deviation).}}
\label{fig:coalitions}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figures/resilience_both}
\caption{{\footnotesize Resilience of the coalition structures when nodes fail randomly (see equation \ref{eq:resilience}) for $ P^{MIN} = 10MW $ (top subplot) and $ P^{MIN} = 80MW $ (bottom subplot)} }
\label{fig:resilience}
\end{figure}

>>>>

{\bf Keywords:} prosumers, smart-grid, coalitions, correlation, stability 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Section I: Introduction
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}

The recent developments in the smart grid area, such as demand side management mechanisms, storage efficiency, or communication architectures and algorithms to name a few, sketch a new power system. The rigid top-down centralized architecture of today’s power grids is expected to evolve in more distributed and flexible systems. Bidirectional power flows, storage, and efficient use of information are some of the key principles that will enable this transition. Since production could be located down to the very end of the distribution networks, nodes that were simply pure loads yesterday could behave tomorrow as generators or loads, further complicating energy usages [5]. 
In this paper, we focus on these "nodes" in the distribution network that can produce and consume electricity. More precisely, we consider a set of agents that own both renewable distributed energy resources (DER) and electrical loads. The production of an agent can be, of course, used to meet its own demand, but in cases when it is over-producing, we consider that it has the possibility to inject, against financial compensation, its extra-production in the grid. Such an agent model is known as a "prosumer" [6]. A key point of this work is to merge the interests of the grid operator with the prosumers in a single utility function. While the latter intend to maximize their benefits with higher production contracts, the operator primary concerns are related to the quality and stability of the injections. The coalitions that we wish to form should thus be both stable and productive. This compromise can be difficult to obtain in real situations, especially when renewable generation and complex spatiotemporal correlation happen.

@@

\begin{figure}
\label{tab:main_notations}
	\begin{tabular}{ | l | r | }
   						N & Number of Prosumers \\
  						$ P_i(t) $ & Extra-production of agent i at time t \\
   						$ P_S^{CRCT} $ & Contract value of coalition S \\
   						$ \phi \in \left[ 0,1 \right] $ & Reliability Threshold \\
   						$ P^{MIN} $ & Minimum Contract value \\
   						$ \alpha $ & Parameter that controls the coalitions sizes \\
   						$ \epsilon $ & Correlation graph filtering threshold \\
   						$ N_{COAL} $ & Number of coalitions \\
   						$ \mathcal{R}_S $ & Resilience of coalition S
 	\end{tabular}
 	\caption{Main notations.}
\end{figure}

@@

\begin{figure}
\label{fig:process}
	\includegraphics[scale=.5]{Fig2}
	\caption{Process.}
\end{figure}

@@

We denote by $ \nu_{i}(t) $ and $ \Psi_{i}(t) $ the wind speed (in $ m.s^{-1} $) and the solar radiance ( in $ W.m^{-2} $) at the location of agent i and at time t, so that :
\begin{equation}
 P_{i}^{P}(t) = \sum_{k \in WT_{i}} \mathcal{F}_{WT}( \nu_{i}(t) ) + \sum_{k \in PV_{i} } \mathcal{F}_{PV}(\Psi_{i}(t) ) 
\end{equation}
Where $ \mathcal{F}_{WT} $ (resp. $ \mathcal{F}_{PV} $) is the power curve for the wind-turbines (resp. photovoltaic panels). We made here the implicit assumption that all wind-turbines (resp. photovoltaic panels) have the same power curve. The model can be easily extended to multiple power curves accounting for different types of generators. More details about power curves and their approximations can be found in the appendix and in \cite{Lydia2014}.

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_1}
\caption{{\footnotesize $ P_{S}^{CRCT \star} $ depending on reliability parameter $ \phi $ for Gaussian distributions (see equation \ref{eq:P_star}). Blue curve with triangles stands for a coalition S with an expected production of 5 units and a standard deviation of 0.5. Under a grid policy of $ \phi = 0.1 $, it is able to announce a contract value of $ P_{S}^{CRCT \star} = 4.36 $. The same coalition in term of expected production ($\mu = 5 $), but with a higher variance ($ \sigma = 1.5 $, cyan curve with circles) can only afford a smaller contract value of $ P_{S}^{CRCT \star} = 3.07 $. The red curve with diamonds stands for a coalition with a higher expected production $(\mu = 7)$, but with a very high unpredictability ($ \sigma = 5 $). For low values of $ \phi $, this coalition is thus heavily penalized and can only afford a contract of $ 0.59 $ units. Under grid policy $ (\phi = 0.1, P^{MIN} = 2) $, this last coalition is thus not allowed to enter the market (red dot below the horizontal dashed line).} }
\label{fig:Gaussian}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_2}
\caption{{\footnotesize Gaussian mean approximation. Subplot \textbf{a} shows how the parameter $ \alpha $ of the utility function should be chosen in function of the mean size of the coalitions (see equation \ref{eq:alpha_star}). Subplot \textbf{b} displays the corresponding utility functions for different values of $ \alpha $. Blue curve with diamonds favors small coalitions of 2 agents while the green one with triangles favors 5 agents coalitions. Finally, the red curve with squares has an optimal size of 15 agents.} }
\label{fig:mean_approx}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_3}
\caption{{\footnotesize Histograms of utility values for coalitions of size 3. Red bars stand for cliques in the decorrelation graph, and blue bars for all the other possible triplets. Cliques tend to exhibit higher utilities than random coalitions.}}
\label{fig:histo_cliques}
\end{figure}

@@

Before running the algorithms, we need to calibrate the utility function by choosing the value of the $ \alpha $ parameter. Recall that the purpose of this parameter is to take into account some constraints on the coalition's sizes if needed. In this paper, we do not have any technical constraints on coalition sizes even if we designed the utility such that these could be taken into account. We select the desired size as being $ \lfloor N/N_{COAL} \rfloor $ (where $ \lfloor.\rfloor $ means floor). Figure \ref{fig:real_utility2} shows how the mean utility of a coalition evolves with its size when the optimum size is set to 40 agents.
\begin{figure}
\includegraphics[scale=.48]{./figs/figure_4}
\caption{{\footnotesize Utility of random coalitions depending on their size. Blue dots show real mean utility values and the thick red curve its smoothed version by applying a Savitzky-Golay filter. On this plot the $ \alpha $ parameter of the utility function was selected according to eq. \ref{eq:alpha_star} in order to favor 40 agents coalitions.}}
\label{fig:real_utility2}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_5}
\caption{{\footnotesize Coalitions formed in the (contract value, volatility) space. The color map indicates the underlying quality. The closer to red the better (high contract values with small volatility). On the opposite, blue areas show poor quality (small contract values with high volatility). Blue dots stand for the decorrelated coalitions that we formed while green squares show correlated coalitions. The smaller yellow markers stand for the gravity centers of the coalition structures. The black dotted line shows how contract values and volatility evolve  when the size of the coalitions increases (a subset of the points are labeled by the size of the coalition they represent). Each point is the average over $ 10^{5} $ unconstrained draws of a random coalition. We also show the distribution of the gravity centers of random coalition structures with red ellipses (center is the mean, and each ellipse corresponds to one  standard deviation).}}
\label{fig:coalitions}
\end{figure}

@@

\begin{figure}
\includegraphics[scale=.48]{./figs/figure_6}
\caption{{\footnotesize Resilience of the coalition structures when nodes fail randomly (see equation \ref{eq:resilience}) for $ P^{MIN} = 10MW $ (top subplot) and $ P^{MIN} = 80MW $ (bottom subplot)} }
\label{fig:resilience}
\end{figure}
